Refactoring the Candidate Cancer Gene Database

### 2021-05-21 presentation.Rmd
furthered presentation content

---
title: "Refactoring the Candidate Cancer Gene Database"
author: "Christopher Tastad"
date: "May 21, 2021"
output:
  revealjs::revealjs_presentation:
    self_contained: false
    center: false
    theme: night
    highlight: zenburn
    transition: none
    reveal_plugins: ["menu", "notes"]
    reveal_options:
      showNotes: "separate-page"
      controls: true
      height: "100%"
      margin: 0.04
      slideNumber: true
      previewLinks: true
---

## Presentation ToC
> - Original Concept
> - Challenges
> - Refactoring the Concept
> - Outcomes of Refactoring
> - Conclusions

@@

@@

@@

## CCGD Overview

@@

## Breakdown of build table process

@@

## OIT setup

@@

# Challenges {.center}

@@

## Forced Upgrades

@@

## Technical Debt
> - test
> - test

@@

::: {#table:projectReqs}
      Feature           Goal                       Framework   Description
  --- ----------------- -------------------------- ----------- -------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  1   server OS         upgrade server OS          RHEL7       Transition of architecture for public server host. This is required by University OIT due to end-of-life schedule for RHEL6.
  2   web front-end     rewrite web interface      Rmarkdown   Improvements to the web interface written in a modern, simplified language. This improved access to content creation and allowed for automation in front-end rendering.
  3   table build       rewrite table build        Rshiny      Rshiny offered a dramatic improvement to replace the existing process by merging the table build back-end with a modern web display of the app interface.
  4   content update    improved admin controls    BASH/R      Old app version confined some content controls to app author, limiting ability to make contributions by product owner (Tim Starr).
  5   version control   implement best practices   git/docs    No version control was used in the original development of the app. This and other documentation practices were expanded in the rewrite.
  6   multiple          resource improvements      codebase    Due to the lack of some best practices, there were many opportunities to make impactful resource improvements.

@@

```{r echo = T}
title: "Candidate Cancer Gene Database"
bibliography: refs/ccgd_paper.bib
nocite: '@*'
output:
    html_document:
        includes:
            in_header: "styles/favicon.html"
            after_body: "styles/footer.html"
        css: styles/styles.css
        theme: readable
---

```{r echo = T}
name: "Candidate Cancer Gene Database"
navbar:
    title: "CCGD"
    left:
        - text: "Home"
          href: index.html
        - text: "Search"
          href: search.html
        - text: "Help"
          href: help.html
        - text: "References"
          href: references.html
        - text: "Contact"
          href: contact.html

@@

@@

## Simplified Codebase
```{r echo = T}
df <- read.csv("ccgd_export.csv") # read in base source file
speciesList <- c("Mouse", "Human", "Rat", "Fly", "Fish", "Yeast")

@@

@@

@@

## Conclusions

@@

.column {
  column-count: 2;
}

>>>>

---
title: "Refactoring the Candidate Cancer Gene Database"
author: "Christopher Tastad"
date: "May 21, 2021"
output:
  revealjs::revealjs_presentation:
    self_contained: false
    center: false
    theme: night
    highlight: zenburn
    transition: none
    reveal_plugins: ["menu", "notes"]
    reveal_options:
      showNotes: "separate-page"
      controls: true
      height: "100%"
      margin: 0.04
      slideNumber: "c/t"
      previewLinks: true
---

## Presentation ToC
> - Original Concept
> - Issue
> - Refactoring the Concept
> - Outcomes of Refactoring
> - Conclusions

@@

## Original Concept

![](./fig/ccgd_header.png)

> - Cancer bio is complicated.
> - Driver genes present primary targets.
> - Screening methods...

<div class="notes">
- look at the origins and maintenance of the ccgd
- why does such a tool exist
- what purpose does it offer the field

- Cancer biology presents a highly complex genetic landscape.
- Cohorts of genes act as drivers for the progression of cancer.
- These genes present targets for study and development of therapeutics.
- Ultimately, we need to understand and simplify cancer gene targets.
- Screening methods offer quicker access to wide swaths of these targets.
</div>

@@

![](./fig/transposon_screen_mech.jpg)

<small>Ranzani, Marco, et al. “Cancer Gene Discovery: Exploiting Insertional Mutagenesis.” Mol Cancer Res, vol. 11, no. 10, 2013, pp. 1141–1158., doi:10.1158/1541-7786.MCR-13-0244.</small>

<div class="notes">
- gene cassette is used to place a genetic payload at some region relative to a transcriptional site to induce mutagenesis
- figure shows 3 mechanisms to derive outcome
- A: enhancer insertion likely upstream of promoter.
- B: vector integrates within some transcriptional range and co-ops promotor function from gene itself.
- C: intragenic insertions may disrupt coding regions generating novel proteins with GoF or LoF outcomes impacting tumorigenesis or tumor suppression.
</div>

@@

![](./fig/transposon_screen.jpg)

<small>Ranzani, Marco, et al. “Cancer Gene Discovery: Exploiting Insertional Mutagenesis.” Mol Cancer Res, vol. 11, no. 10, 2013, pp. 1141–1158., doi:10.1158/1541-7786.MCR-13-0244.</small>

<div class="notes">
- transposon method is applied en masse within mice.
- process of extraction, isolation, seq, and analysis identifies CIS groups.
- validation takes place in vitro or in vivo for oncogenic potential with orthologous association to human candidates.
- idea for CCGD comes in at the star.
</div>

@@

## CCGD Goals

> To facilitate analysis of driver genes to catalog all common insertion sites (CISs) and their corresponding genes identified in published studies using transposon insertional mutagenesis.

> - Create an interactive web tool to quickly refine the stated data set by a variety of parameters.
> - Draw from universal resources to define homology relationships for candidate genes between animal models.
> - Curate disease classification and meta-analysis through complete CIS listing and relative rank and strength scores.
> - Offer easy access to external tools from filtered results for deeper biological context.

<div class="notes">
- relative rank was based either on the number of insertions in a given CIS or the P-value associated with the given CIS
- The relative rank was calculated by ranking all CISs in a given study and then assigning grades
- We assigned a letter grade of A for the top 10%, B for the top 11–25%, C for the top 26–50% and D for the bottom 50%.
</div>

@@

## CCGD Implementation

**ETL Process**

scrape data &rarr; homology association &rarr; integrate into table

![](./fig/search.png)

<div class="container">

<div class="col">
#### Data Sources/Links
- NCBI Gene
- NCBI HomoloGene
- NCBI PubMed
- Sanger COSMIC
- Sanger CGC
- GeneCards
</div>

<div class="col">
> - Filtered results are downloadable.
> - Wrapped in a simple, instructional website built with classic LAMP stack.
> - Hosted by UMN-OIT.
</div>

<div class="notes">
- Search Within - mouse human rat fly fish yeast
</div>

@@

# Issue {.center}

@@

## Issue

### Forced OS upgrade from UMN-OIT host

@@

## Challenges

#### Can this application be upgraded as-is?

<div class="notes">
- several of these items blend limitations that are dependency-based, while others amount to an inconvenience.
</div>

**Primary Challenges**

> - Machine displacement in provisioning by UMN-OIT
> - Use of abandonded, proprietary widgets
> - Outsized SQL database

**Secondary Challenges**

> - Documentation missing visibility for admins
> - Total app disk use
> - Lack of version control
> - Odd hosting paradigm from UMN-OIT
> - Disorderly codebase, residual WYSIWYG effects
> - Lack of modularity in back-end
> - Pending web standards compliance issues

<div class="notes">
- these are a combination of both limitations and potential goals
</div>

## Challenges

#### Can this application be upgraded as-is?

**Primary Challenges**

- <span style="color: #a3be8c;">**Machine displacement in provisioning by UMN-OIT**</span>
- Use of abandonded, proprietary widgets
- <span style="color: #a3be8c;">**Outsized SQL database**</span>

**Secondary Challenges**

- Documentation missing visibility for admins
- Total app disk use
- Lack of version control
- Odd hosting paradigm from UMN-OIT
- Disorderly codebase, residual WYSIWYG effects
- Lack of modularity in back-end
- Pending web standards compliance issues

## Challenges

### Ultimately, a drop-in transition was not possible.

@@

::: {#table:projectReqs}
      Feature           Goal                       Framework   Description
  --- ----------------- -------------------------- ----------- -------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  1   server OS         upgrade server OS          RHEL7       Transition of architecture for public server host. This is required by University OIT due to end-of-life schedule for RHEL6.
  2   web front-end     rewrite web interface      Rmarkdown   Improvements to the web interface written in a modern, simplified language. This improved access to content creation and allowed for automation in front-end rendering.
  3   table build       rewrite table build        Rshiny      Rshiny offered a dramatic improvement to replace the existing process by merging the table build back-end with a modern web display of the app interface.
  4   content update    improved admin controls    BASH/R      Old app version confined some content controls to app author, limiting ability to make contributions by product owner.
  5   version control   implement best practices   git/docs    No version control was used in the original development of the app. This and other documentation practices were expanded in the rewrite.
  6   multiple          resource improvements      codebase    Due to the lack of some best practices, there were many opportunities to make impactful resource improvements.

@@

<div class="container">

<div class="col">
```{r echo = T}
---
title: "Candidate Cancer Gene Database"
bibliography: refs/ccgd_paper.bib
nocite: '@*'
output:
    html_document:
        includes:
            in_header: "styles/favicon.html"
            after_body: "styles/footer.html"
        css: styles/styles.css
        theme: readable
---

```
</div>

<div class="col">
```{r echo = T}
---
name: "Candidate Cancer Gene Database"
navbar:
    title: "CCGD"
    left:
        - text: "Home"
          href: index.html
        - text: "Search"
          href: search.html
        - text: "Help"
          href: help.html
        - text: "References"
          href: references.html
        - text: "Contact"
          href: contact.html
---
```
</div>

@@

<div class="notes">
- ADD SOME NOTES
</div>

@@

<div class="notes">
- codebase simplified in a few ways.
- build table process was dramatically streamlined.
</div>

```{r echo = T}
export <- homogTable %>%
  full_join(df) %>%
  group_by(MouseId) %>%
  distinct(Study, .keep_all = T) %>%
  add_count(name = "Studies") %>%
  ungroup() %>%
  mutate(CGC = HumanName %in% cgc) %>%
  mutate(COSMIC = HumanName %in% cosmic)

write.csv(export, file = "../table_app/ccgd_export.csv", row.names = F)

deployApp(appDir = "../table_app", forceUpdate = T, launch.browser = F)
```

#### build_table.R table write and app deployment snippet

<div class="notes">
- shiny snippet represents full set of code needed to render app ui. app is 200 lines total.
- lines of code replacement is orders of magnitude. possible 1:100.
- use of this script allowed for elimination of entire sql database.
</div>

```{r echo = T}
export <- homogTable %>%
  full_join(df) %>%
  group_by(MouseId) %>%
  distinct(Study, .keep_all = T) %>%
  add_count(name = "Studies") %>%
  ungroup() %>%
  mutate(CGC = HumanName %in% cgc) %>%
  mutate(COSMIC = HumanName %in% cosmic)

write.csv(export, file = "../table_app/ccgd_export.csv", row.names = F)

deployApp(appDir = "../table_app", forceUpdate = T, launch.browser = F)
```

#### build_table.R table write and app deployment snippet

```{r echo = T}
df <- read.csv("ccgd_export.csv") # read in base source file
speciesList <- c("Mouse", "Human", "Rat", "Fly", "Fish", "Yeast")

@@

<div class="notes">
- Page content is now written in a highly accessible syntax and rendered with a lean script.
</div>

@@

```{r echo = T}
setwd("..")

# render core webpages
rmarkdown::render("index.Rmd")
rmarkdown::render("search.Rmd")
rmarkdown::render("help.Rmd")
rmarkdown::render("references.Rmd")
rmarkdown::render("contact.Rmd")
rmarkdown::render_site("index.Rmd")
```

#### knit_site.R page and site rendering script

@@

## Conclusions
- Stated requirements set out by this project have been met.
- A new, more modern and simplified framework has been implemented across the full application stack.
- All original functionality has been sustained.
- Design and engineering considerations were given to ease future maintenance.

## Future Directions
- Continuous integration/continuous deployment
- Anticipating future upgrades and framework evolution

@@

.container{
    display: flex;
}

.col{
    flex: 1;
}
